**Generalized additive models to analyze biomedical non-linear longitudinal data in R: Beyond repeated measures ANOVA and Linear Mixed Models**

### 2021-11-11 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/01-Background.Rmd
Updated

During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data. The recognition on the shortcomings of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data [@skala2010;@vishwanath2009]. Briefly, LMEMs incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age). When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006]. However, LMEMs impose restrictions in the distribution of the errors  of the random effects, which need to be normally distributed and independent [@gueorguieva2004;@barr2013]. And even more importantly, LMEMs also assume by default a linear relationship between the response and time [@pinheiro2006], making them unsuitable to analyze data non-linear with non-linear trends (polynomial effects can be used with LMEMs, but this approach has its own shortcomings as we discuss in Section {#GAM-theory}) .

>>>>

During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data. The recognition on the shortcomings of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data [@skala2010;@vishwanath2009]. Briefly, LMEMs incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age). When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006]. However, LMEMs impose restrictions in the distribution of the random effects, which need to be independent [@gueorguieva2004;@barr2013]. And even more importantly, LMEMs also assume by default a linear relationship between the response and time [@pinheiro2006] (polynomial effects can be used with LMEMs, but this approach has its own shortcomings as we discuss in Section {#GAM-theory}) .

### 2021-11-11 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
Updated

>>>>

# The contemporary statistical landscape

Although the limitations of rm-ANOVA and LMEMs to analyze non-linear trends in longitudinal data have been presented in a general way in the previous section, a proper understanding of the matter would be incomplete without a brief description of the current accepted knowledge regarding statistical models. The advances over the last two decades in computational power and the theoretical framework for statistical models have significantly changed the hierarchy of statistical models. Because the topic itself is far outside of the scope of this work

## Linear Models (LMs) and Linear Mixed Models (LMEMs)

Linear models (LMs) are those that assume a normal (Gaussian) distribution, such as ANOVA. These are by far the models most commonly used to analyze data within the biomedical research community, and only incorporate _fixed effects_. On the other hand, Linear Mixed Effect Models (LMEMs) are those that still assume normality but also incorporate _random effects_, as it has been described before.

## Generalized Linear Models (GLMs)

A major limitation of LMs is their assumption of normalit, and although it can be forced by the use of a transformation, this procedure can lead to poor model performance [@ohara2010], and can cause problems with the biological interpretation of the model estimates. McCullagh and Nelder [@nelder1972] defined General Linear Models (GLMs) as an extension of LMs, where the data does not need to be normally distributed. To achieve this, consider the following equations:

\begin{equation}
y_{ijt} \sim \mathcal{D}(\mu_{ijt},\phi)
(\#eq:GLM-y)
\end{equation}

Where $y_{ijt}$ is the observed response, which is assumed to follow some distribution $\mathcal{D}$ with some mean $\mu_{ijt}$ and a dispersion parameter $\phi$. The mean $\mu_{ijt}$ is also known as the _expected value_ ($E$) of the observed response.

\begin{equation}
E(y_{ijt})=\mu_{ijt}=g^{-1}(\eta_{ijt})
(\#eq:GLM-Expectation)

In contemporary Statistics,

Need the two first equations

and this one

\begin{equation}
$\eta_{ijt}=\beta_0+\beta_1x_{ijt}$ 
(\#eq:GLM-eta)
\end{equation}

## Generalized linear mixed models (GLMMs)

$\eta_{ijt}=\beta_0+\beta_1x_{ijt}+\mu_{ij}$ GLMM
